---
title: 'Lab 1 Inclass: Review/Introduction to R and Data Structures'
author: '[PUT YOUR NAME HERE]'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
subtitle: 'SOC 325: Quantified-Self'
---

```{r, include=FALSE}
library(tidyverse)
#install.packages("coronavirus")
```

# 1: Getting started: Editing and running code

There are several ways we can run code:

  - Highlight lines in the editor window and click **Run** at the top or hit `Ctrl` + `Enter` or `CMD` + `Enter` (Mac) to run them all.
  - With your caret on a line you want to run, hit `Ctrl` + `Enter` or `CMD` + `Enter` (Mac). Note your caret moves to the next line, so you can run code sequentially with repeated presses. 
  - Type individual lines in the console and press enter. 
  - In R Markdown documents, click within a code chunk and click the green arrow to run the chunk. The button beside that runs all prior chunks. 
  
**Question 1.1: run the following lines of code.**

```{r}
123 + 456 + 789
sqrt(400)
```

Now run the following code in console: (11-2. This code obviously is incomplete. What do you see in your console?

You may notice that here, R is used as a calculator. Now, follow the instructions on this website [R as a calculator](https://sphweb.bumc.bu.edu/otlt/MPH-Modules/BS/R/R1_GettingStarted/R1_GettingStarted3.html) and play out with calculation in the console. Write a brief summary of your experience below:

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}

```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

  
# 2: Functions and help

`sqrt()` is an example of a function in R. 

If we didn't have a good guess as to what `sqrt()` will do, we can type `?sqrt` in the console and look at the **Help** panel on the right. 

**Question 2.1: looking at the Help file of function sqrt().**

**Argument** are inputs to a function. In this case, the only argument to `sqrt()` is x which can be a number or a vector of numbers.

Help files provide documentation on how to use functions and what functions produce. 

Note, that the R-chunk is turned off as help doesn't work in R Markdown. Write out the code to see the help file below:

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r,eval=FALSE}

```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

# 3: Objects

R stores everything as an object, including data, functions, models, and output. 

Creating an object can be done using the assignment operator: `<-`
When you're typing inside of a code chunk (that's where the code goes), you can use `alt` plus `the minus key` to quickly type the arrow. Assigen 5 to "a".

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}

```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

**Question 3.1: Create an new object called "new_object". Assign value 144 to it.**

```{r}
new_object <- 144
```
Operators like `<-` are functions that look like symbols but typically sit between their arguments (e.g. numbers or objects) instead of having them inside () like in `sqrt()`.

You can display or "call" an object simply by using its name:
```{r}
new_object
```

An object's name represents the information stored in that object, so you can treat the object's name as if it were the values stored inside. 
```{r}
new_object + 10
```

```{r}
new_object + new_object
```

Next, we are going to create a vector. A vector is a series of elements, such as numbers. You can create a vector and store it as an object in the same way. To do this, use the function `c()` which stands for "combine" or "concatenate".

```{r}
new_object <- c(4,9,16,25,36)
new_object
```

If you name an object the same name as an existing object, it will overwrite it. Overwrite "new_object" with "Hello World".

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}

```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

**Question 3.2: calculate the square root of vector `new_object`.**

The same principles can be used to create more complex objects like matrices, arrays, lists and dataframes (lists of equal-length vectors). Each element of the dataframe can be thought of as a column, and the length of each element is the number of rows. 

Most data sets we work with will be read into R and stored as a dataframe, so the labs will mainly focus on working with these objects. 

Let's take a look at an example of data on covid-19. To do so, we will first have to install a package called "coronavirus", developed by the Johns Hopkins University. Packages contain premade functions and/or data we can use. It is R's strength that there are a wide variety of packages! **Note, we wont be doing any installation on R, as this is done through UW IT, but if you install R on your own computer you will need to be able to install packages.**

```{r}
#install.packages("coronavirus") # only install once, don't re-install until you update R
#library(coronavirus)
```

The package name here is a search term, i.e.text, not an object, so we must use quotation marks. 

We then use command `library()` to load the package, and `data()` command to take a look at the built-in dataset. To take a deeper look at the package, visit [Package 'coronavirus'](https://cran.r-project.org/web/packages/coronavirus/coronavirus.pdf).

```{r}
library(coronavirus)
data(coronavirus)
```

This dataset contains the daily summary of Coronavirus cases (confirmed, death, and recovered), by state/province. Let's view the dataset.

Print out the first ten rows of the "coronavirus" data set using the `head()` function. `head` shows the first 6 rows by default. Change the following code to show the first 10 rows.** 

```{r}
head(data.frame(num=1:26,let=letters,LET=LETTERS), n = 6)
```

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}

```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

# 4: Basics about R Markdown

This is a R Markdown file. The Markdown file enables document analyses by combining text, code and output, so that you don't have to copy and paste into word. You can control the amount of code to display in your file, and it is easy for collaborators to understand. 

R Markdown is able to produce various document types such as HTML, PDF, Word, Powerpoint and Presentation slides. It also works with LaTex and HTML for mathematics and other format control. 

Let's first look at the header of an .Rmd file, a [YAML](http://yaml.org) code block. 

Inside R Markdown, R code are written into chunks. Code is sandwiched between sets of three backticks and `{r}`. Below is a chunk of code:

```{r}
# point R to the working directory on your computer
#setwd() ## we don't need to do this on the server
```

You can make a new code chunk by using the insert menu above, typing the backtics by hand, or by pressing `CMD`+`option`+`i` on a Mac or `CTRL` + `ALT` + `i` on windows or linux.

Chunks have options that control what happens with their code, such as:

* `echo=FALSE`: Keeps R code from being shown in the document. 
* `eval=FALSE`: Shows R code in the document without running it. 
* `include=FALSE`: Hides all output but still runs code.
* `results='hide'`: Hides R's (non-plot) output from the document
* `cache=TRUE`: Saves results of running that chunk so if it takes a while, you won't have to re-run it each time you re-knit the document.
* `fig.height=5`, `fig.width=5`: modify the dimensions of any plots that are generated in the chunk (units are in inches). 

**Question 4.1: hide outputs in the chunk where we loaded package.**

Sometimes we want to combine text with values. We do that using code in single backticks starting off with `r`. 

Four score and seven years ago is the same as `r 4*20+7` years. Notice that R could also be used as a calculator.

We are able to reference an object we already saved in a chunk:
```{r}
x <- sqrt(77)
```

The value of `x` rounded to the nearest two decimals is `r round(x,2)`. 

Inserting in-line code with R helps you prevent silly mistakes and makes your coding much more efficient - you don't have to manually update your hard-coded number if the calculation changes. 

<span style="color: red;">YOUR SOLUTION HERE</span>

Insert a line of code that produces `x` + `15` = YOURCODEHERE

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

**Question 4.2: mix in-line R calculations/references with text. You may not hard-code any numbers referenced.**

Start your text here:

More about RMarkdown: [R Markdown Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).

**Question 4.3: Taking a look at the cheat sheet, and try to work on the following outputs.**

<span style="color: red;">YOUR SOLUTION HERE</span>

Italicize this line. 

Make this line bold. 

Strike through this line. 

Create a **block code** of the following poem: "No man is an island, Entire of itself, Every man is a piece of the continent,
A part of the main."

Finally, how do we take a look at the final output of our Markdown document?

Click on **Knit**, select *Knit to HTML*. 

<span style="color: red;">END OF YOUR SOLUTION HERE</span>


[Managing Data Frames](http://uc-r.github.io/dataframes)

# 5: Opening this file

All of the labs will be available on JupyterHub and you should start by loading `Lab 1.Rproj`. Write up your experience loading up the `Rproj` file. List three advantages of using the `RProj`.


<span style="color: red;">YOUR SOLUTION HERE</span>


<span style="color: red;">END OF YOUR SOLUTION HERE</span>

# 6: Describe data sources 

Besides a package, the easiest way to work with external data is for it to be stored in a *delimited* text file, e.g. comma-separated values (.csv) or tab-separated values (.tsv). 

In the same directory as this `.Rmd` file and the project file, there's a directory called `data`. Inside, there's a csv file called `seattle_airbnb.csv`. This contains data about 100 Airbnb listings from Seattle. 

These data come from Inside Airbnb, http://insideairbnb.com/. Go to the website and have a look at the About, Behind, and Get the Data pages. *Use what you read to answer the questions below. Just type your answers below the questions.*

**Question 6.1: What was the _context_ of this data's production, ie:** 
  
  <span style="color: red;">YOUR SOLUTION HERE</span>

   - Who created this data set?
   - How did they do it?
   - Where did the data come from?
   - Is this active or passive data collection?

<span style="color: red;">END OF YOUR SOLUTION HERE</span>
  

# 7: Set up and read data into R

The package we'll use today is called "tidyverse." It's a collection of packages for data manipulation, exploration, and visualization.  

You can read more about the tidyverse here: https://www.tidyverse.org/

Follow the instructions we learned  to load the package `tidyverse`.


<span style="color: red;">YOUR SOLUTION HERE</span>
```{r}

```
<span style="color: red;">END OF YOUR SOLUTION HERE</span>

To use data inside R, we first have to import, or *read*, that data into our environment. The chunk below reads the example data we'll use for this module.

```{r}
airbnb_data <- read_csv("data/seattle_airbnb.csv")
```

Notice that here, we create an object of a dataframe called "airbnb_data". Print out the first ten rows.

<span style="color: red;">YOUR SOLUTION HERE</span>
```{r}

```
<span style="color: red;">END OF YOUR SOLUTION HERE</span>

**Question 8: What is the role of each component in the above line of code?**

 - airbnb_data: the variable name we used for the data frame
 - `<-` :
 - read_csv() :
 - "data/seattle_airbnb.csv" 
 
 
# 8: Looking at the data

Let's take a look at the data.   

You can also look at the entire data set using RStudio's built-in viewer. To use that, we use the function 'View().' We can run that command from the console, or from a code chunk:

```{r}
# "un-comment" the line below this one, by removing the '#' and the space
# View(airbnb_data)
```

What if you want to look at the *last* several rows of a data frame instead of the first several rows?

Let's read the documentation for `head` by typing `?head` into the console.

**Question 8.1 Based on what you found out, show the last 5 rows of airbnb_data**

<span style="color: red;">YOUR SOLUTION HERE</span>
```{r}

```
<span style="color: red;">END OF YOUR SOLUTION HERE</span>

You can extract a single column by name using `$`. Type the name of the dataframe (airbnb) first, then `$` and finally the name of the column. 

**Question 8.2 Use `$` to display the 'price' column**

<span style="color: red;">YOUR SOLUTION HERE</span>
```{r}

```
<span style="color: red;">END OF YOUR SOLUTION HERE</span>

# 9. Subsetting data

Sometimes we have a large dataset but we only need to work with a subset of it. There are several ways to modify and subset a dataframe. Here, we will learn to subset with indices. 

**Question 9.1: Change the code below so that we load the haven package and make a new object called `twitter_survey` by using the read_sav() function to read the file "data/Pew_Twitter_Study_for_release.sav".**

```{r}
library(tidyverse)
library(haven)
airbnb_data <- read_csv("data/seattle_airbnb.csv")
```

Twitter data
```{r}
twitter_survey<-read_sav("data/Pew_Twitter_Study_for_release.sav")
```

To get a quick look at names of the variables in this data frame, run the below code:
```{r}
names(twitter_survey)
```

There are many variables but we only need three of them: CaseID, TWITTER_USE, VOTED. To manipulate data frames in R, we can use the `[]` notation to access the indices for the observations and the variables. It is easiest to think of the data frame as a rectangle of data where the rows are the observations and the columns are the variables. The indices for a rectangle of data follow the RxC principle; in other words, the first index is for Rows and the second index is for Columns [R, C]. When we only want to subset variables (or columns) we use the second index and leave the first index blank. Leaving an index blank indicates that you want to keep all the elements in that dimension.

```{r}
twitter_subset <- twitter_survey[,c(1,9,15)]
```

If the variables we want are in consecutive columns, we can use the colon notation rather than list them using the c function.

```{r}
twitter_subset2 <- twitter_survey[,1:4]
```

**Question 9.2: Create a subset of dataframe twitter_subject, taking the first 100 rows and variables PARTY and PPWORK. Assign this object with name "my_twitter_survey".**


<span style="color: red;">YOUR SOLUTION HERE</span>
```{r}

```
<span style="color: red;">END OF YOUR SOLUTION HERE</span>



# References
Cuture, Victor, Jonathan I. Dingel, Allison Green, Jessie Handbury, and   Kevin R. Williams. 2021. “JUE Insight: Measuring Movement and Social    Contact with Smartphone Data: A Real-Time Application to COVID-19.”       Journal of Urban Economics 103328. doi: 10.1016/j.jue.2021.103328.

Charles Lanfear, [Introduction to R for Social Scientists](https://clanfear.github.io/CSSS508/)

[Short Classic Poems](https://medium.com/@EmEmbarty/31-of-the-best-and-most-famous-short-classic-poems-of-all-time-e445986e6df)


  