---
title: "SOC 325 Lab 5 - Analyzing Watch Data"
subtitle: "Soc 325: Quantified-Self"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---


# Pyschology of R, RStudio and Labs

Scripting is HARD. That is ok, many hard things are useful! This is very helpful if you want to work as a data scientist. You need to familiarize yourself with the practices of coding and building analysis scripts.

**Why**

* Your work has to be reproducible 
* Your work has to be able to hand off to someone else
* You need to be able to do lots of different things

**You Don't need to do everything from scratch!!**

* It is ok to borrow code, use [stack overflow](https://stackoverflow.com/), [google](https://google.com), etc.
* Use books like [R for Data Science](https://r4ds.had.co.nz/introduction.html).

**Strategies for working in RMarkdown**

* DO **NOT** code a bunch and then *knit* -- This will almost assuradely not work and crash!
* Do make small changes and *knit*
  + If you have a part that takes a long time (e.g. reading in data) turn `cache=TRUE` on in the R Chunk. This will pre-load it and make it faster.


# Analysis of Sleep Watches

## Load Libraries Needed For this Lab

For library management we use use the library `pacman`. The key with pacman is it will install and load the packages if needed. This makes our life easier!

The first line of code is `if (!require("pacman")) install.packages("pacman")` to install pacman if needed.

The second line `pacman::p_load([PACKAGES HERE])` uses `pacman` to load and install all the packages for the lab.


```{r,message=FALSE,warning=FALSE}
## Run this code, to manage packages and install as needed.
# Install pacman
if (!require("pacman")) install.packages("pacman", repos = "http://cran.us.r-project.org")
# p_load function loads packages if installed, or install then loads otherwise
pacman::p_load(tidyverse,haven,lubridate,ggthemes,kableExtra,here,ggpubr,rstatix,Routliers,reshape2,Routliers,gt,ggforce,car,data.table,zoom,magrittr,zoo,infer)
```

# Analyzing data from sleep watches

This is the data from 

* Dunster, Gideon P., Isabelle Hua, Alex Grahe, Jason G. Fleischer, Satchidananda Panda, Kenneth P. Wright Jr, Céline Vetter, Jennifer H. Doherty, and Horacio O. de la Iglesia. "Daytime light exposure is a strong predictor of seasonal variation in sleep and circadian timing of university students." Journal of Pineal Research (2022): e12843

The article is part of our weekly reading and available [here](https://drive.google.com/file/d/1RM2rgl2n8cyv77_BdoQAbycW7TWAMCx4/view?usp=share_link). *I highly recommened reading it!*

We will look at daily and weekend/weekday variation:

## Violin Plots

```{r,message=FALSE,warning=FALSE}
df <- read.csv (here("data","IndividualSleepSummary.csv"))

#Set year and subject id as factors
df$year <- as.factor (df$year)
df$subjectid <- as.factor(df$subjectid)


####Violin plots (plots days separately) #########
df$onset_mean <- df$onset_mean - 24

#Clock time axis
yaxis <- c("18", "20", "22", "00", "02", "04", "06", "08", "10", "12", "14", "16")
yaxis <- str_pad(yaxis, 2, pad = "0")

#Plot
ggplot(df, aes(x = year, y = onset_mean)) + #Change y axis to whatever plotting (duration, onset, etc.)
  geom_violin(aes(color = year), trim = FALSE, show.legend = F,
              position = position_dodge(width = 0.3)) +
  stat_summary(aes(x = year, y = onset_mean, color = year), fun = "mean", geom = "point",
               size = 8, show.legend = F) + #Change y axis to whatever plotting (duration, onset, etc.)
  geom_sina (aes(fill = subjectid, group = year),
             show.legend = F, shape = 21,
             size = 3.5, alpha = 0.3, maxwidth = .05, 
             position = position_dodge(width = 0.5)) +
  theme_classic () +
  theme(axis.text.x = element_text(size = 25, color = "black"),
        axis.text.y = element_text(size = 25, color = "black"),
        axis.title.y = element_text(size = 25, color = "black", face = "bold" ),
        legend.position = "none",
        strip.text = element_text(size = 25),
        plot.title = element_text(hjust = 0.5, size = 28, face = "bold")) +
  scale_color_manual (values = c("#009E73", "#0072B2", "#D55E00")) + #Change colors to match other plots
  labs(title = "Sleep Duration") +
  ylab ("Clock Time") + #y-axis label, change for duration (Hours)
  xlab ("") + #x-axis label
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  #Set y- axis breaks below to be the same as in y-axis vector
  #will need to change for each parameter
  #clock time hour labels not needed for duaration plot
  scale_y_continuous(limits = c(-6, 16), breaks = seq (-6, 16, 2) , labels = yaxis) +
  facet_wrap(~day)
```

## Waveform Plots

```{r,message=FALSE,warning=FALSE}
filenames <- list.files(path=here("data"),pattern = "\\.csv$") #Get .csv file names in directory
filenames<-filenames[-1]


#Combine data frame with all .csv files data
df<- lapply(here("data/",filenames), fread) %>% 
  set_names(filenames) %>% 
  rbindlist(idcol = "subjectid") #set column with identifier (subject id)

df$subjectid <- substr(df$subjectid,8,9) #make subjectid column into correct number

#Set subject ID as correct year
#If error "column 1 of item 22 does not match with column 1 of item 1"
#Go into file (in excel) named and format data the same as other files
df <- df %>% mutate(year =
                      case_when(subjectid <= 10 ~ "2019", 
                                subjectid >=25 & subjectid <=49 ~ "2020",
                                subjectid >= 50 ~ "2021"))

#make subject, year, and day a factor
df$subjectid <- as.factor (df$subjectid)
df$day <- as.factor (df$day)
df$year <- as.factor(df$year)

#Make time into date/time format
df$Time <- as.POSIXct(df$Time, format="%H:%M:%S", tz = "America/Los_Angeles" )
df$Time <- strftime(df$Time, format="%H:%M")

#Alter data for running average #####

#Log of data (activity and light) + 1 - to not have N/A for zeros (for every individual file)
df$White.Light <- log(df$White.Light+1)

#rolling average for activity in light added to columns (grouped by subject)
df <- df %>%
  group_by(subjectid) %>%
  dplyr::mutate(meanactivity = rollmean(Activity, k = 6, fill = NA),
                loglight = rollmean(White.Light , k = 6, fill = NA))

#Mean for each time point seperated by subject (one point for each subject at each timepoint) ####
activity <- df %>%
  group_by(subjectid, Time, day, year) %>%
  dplyr::summarize(activity = mean(meanactivity, na.rm=TRUE)) #School day activity

light <- df %>%
  group_by(subjectid, Time, day, year) %>%
  dplyr::summarize(light = mean(loglight, na.rm=TRUE)) #weekend activity

#Mean for each timepoint seperated by day (one point per timepoint) ####
#Change time to a factor
activity$Time <- as.factor(activity$Time)
light$Time <- as.factor(light$Time)

#Change format of dataframe for saving file ####

activity_file <- activity %>%
  pivot_wider(names_from = subjectid, values_from = activity)

light_file <- light %>%
  pivot_wider(names_from = subjectid, values_from = light)


#Summarize activity data
activityplot <- activity %>%
  group_by(Time, year, day) %>%
  get_summary_stats(activity, type = "mean_se")

#summarize light data
lightplot <- light %>%
  group_by(Time, year, day) %>%
  get_summary_stats(light, type = "mean_se")

#Alter data for plotting #####
#Change Time column factor levels, so graph is plotted 16:00pm - 15:50pm
#Using row numnbers
activityplot$Time <- factor(activityplot$Time, levels(activityplot$Time)[c(97:144,1:96)])
lightplot$Time <- factor(lightplot$Time, levels(lightplot$Time)[c(97:144,1:96)])

level <- as.data.frame(levels (lightplot$Time)) #check that levels are correct

#Make waveform plot for DAY comparisons (School day v. weekend) ####
#Create wave forms
lightplot %>% #set correct data (light or activity)
  filter(year == "2019") %>% #filter for year
  ggplot(aes(x = Time, y = mean, group = day)) +
  geom_line(aes(color = day), lwd = 1) + #Add line
  geom_pointrange(aes(ymin=mean-se, ymax=mean+se, color = day), size = 0.5) +
  theme_classic() +
  xlab("Clock Time") +
  ylab("Illuminance (log [lux])") + #Axis title set for illuminance or activity
  ylim (0,4) + #y axis scale
  theme(axis.text.x = element_text(angle = 50, hjust = 1), 
        legend.title = element_blank(), legend.position=c(.88, .3), #remove legend title & adjust position on graph
        legend.text=element_text(size=18), #legend text size
        axis.text=element_text(size=18, color = "black"),
        axis.title=element_text(size=18,face="bold", color = "black"),
        #legend.position="none",
        strip.text = element_text(size = 18),
        plot.title = element_text(hjust = 0.5, size = 22, face = "bold")) +
  scale_color_manual(values=c("#E69F00", "#56B4E9")) + #colors for School day vs weekend
  labs (title = "2019") +
  scale_x_discrete(breaks = levels(activityplot$Time)[c(T, rep(F, 5))]) #X axis


#Make waveform plot for year comparisons (2019 v. 2020 v. 2021) ####
lightplot %>% #set correct data (light or activity)
  filter(day == "Weekday") %>% #filter for day (school day or weekend)
  ggplot(aes(x = Time, y = mean, group = year)) +
  geom_line(aes(color = year), lwd = 2) + #Add line
  geom_pointrange(aes(ymin=mean-se, ymax=mean+se, color = year), size = 0.5) +
  theme_classic() +
  scale_x_discrete(breaks = levels(lightplot$Time)[c(T, rep(F, 5))]) + #X axis
  xlab("Clock Time") +
  ylab("Illuminance (log[lux])") + #Axis set for illuminance or activity
  ylim (0,4) +
  labs (title = "Weekday") +
  theme(axis.text.x = element_text(angle = 50, hjust = 1), 
        legend.title = element_blank(), legend.position=c(.88, .3), #remove legend title & adjust position on graph
        legend.text=element_text(size=23), #legend text size
        axis.text=element_text(size=21, color = "black"),
        axis.title=element_text(size=21,face="bold", color = "black"),
        plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
        #legend.position = "none"
        ) +
  scale_colour_manual(labels=c("2019", "2020", "2021"), 
                      values=c("#009E73", "#0072B2", "#D55E00")) #color for years

```

# Assignment

For this lab we are going to analyze sleep data from a Fitbit tracker from Kaggle - [data](https://www.kaggle.com/datasets/riinuanslan/sleep-data-from-fitbit-tracker).

We can load this data directly,

```{r}
November_sleep<- read_csv(here("fitbit_data","November Sleep Data - Sheet1.csv"))%>%na.omit()
December_sleep<- read_csv(here("fitbit_data","December Sleep data - Sheet1.csv"))%>%na.omit()
January_sleep<- read_csv(here("fitbit_data","January sleep data - Sheet1.csv"))%>%na.omit()
February_sleep<- read_csv(here("fitbit_data","February sleep data - Sheet1 (1).csv"))%>%na.omit()
March_sleep<- read_csv(here("fitbit_data","March sleep data - Sheet1.csv"))%>%na.omit()
April_sleep<- read_csv(here("fitbit_data","April sleep data - Sheet1.csv"))%>%na.omit()

```

### Example analysis

Let's look at November


**Step 1** 

Clean data

```{r}
## Clean up the data
nov_clean<-November_sleep%>%mutate(
    st = str_split(`SLEEP TIME`," - ",simplify=TRUE)[, 1],
    wt= str_split(`SLEEP TIME`," - ",simplify=TRUE)[, 2],
    sd=if_else(str_detect(st,"am"),as.Date(DATE,"%m/%d/%Y")+1,as.Date(DATE,"%m/%d/%Y")),
    sleep_time = as.POSIXct(paste(sd,
            st),
            format='%Y-%m-%d %I:%M %p'),
    wd=if_else(str_detect(wt,"am"),as.Date(DATE,"%m/%d/%Y")+1,as.Date(DATE,"%m/%d/%Y")),
    wake_time = as.POSIXct(paste(wd,
            wt),
            format='%Y-%m-%d %I:%M %p'),
    time_slept_min= difftime(wake_time,sleep_time,units="mins"),
    time_slept_hour= difftime(wake_time,sleep_time,units="hour")
)
```

**Step 1.5: Print out Data**

```{r}
nov_clean%>%kbl(digits = 2) %>%
  kable_styling()%>%
   kable_paper("hover", full_width = F)
```

**Step 2: Plot data**

```{r}
## Check histogram of the data
nov_clean%>%
  ggplot(aes(x=as.numeric(time_slept_hour)))+
  geom_histogram()+
  theme_bw()

## Check Scatter plot
nov_clean%>%
  ggplot(aes(y=as.numeric(time_slept_hour),x=`SLEEP SCORE`,color=as.factor(NOVEMBER)))+
  geom_point()+
  theme_bw()

## Check linear relationship between time slept and sleep scor
nov_clean%>%
  ggplot(aes(y=as.numeric(time_slept_hour),x=`SLEEP SCORE`))+
  geom_point()+
  geom_smooth(method="lm",formula='y ~ x')+
  theme_bw()

## Check linear relationship between time slept and sleep score
## Log-Log plot

nov_clean%>%
  ggplot(aes(y=as.numeric(time_slept_hour),x=`SLEEP SCORE`))+
  geom_point()+
  scale_y_log10()+
  scale_x_log10()+
  geom_smooth(method="lm",formula='y ~ x')+
  theme_bw()
```

**Step 3: Explore the data**

```{r}
## Check that it makes sense

## Check Correlations between Sleep Score and Hourse Slept
nov_clean%>%summarise(cor=cor(`SLEEP SCORE`,as.numeric(time_slept_hour),use="na.or.complete"))

## Check the linear relationship between Day of the week and 
#### sleep score
summary(lm(`SLEEP SCORE`~as.character(NOVEMBER)-1,data=nov_clean))

## Check the linear relationship between Day of the week and 
#### hours slept
summary(lm(as.numeric(time_slept_hour)~as.character(NOVEMBER)-1,data=nov_clean))
```


## Question 1

In this question we will look at December and January Sleep

### Questions 1.1

Combine December and January data using the tidyverse method `OBJECT1 %>% bind_rows(OBJECT2)`. Save this object as `dec_jan`.

**Hint From Lecture**


```
## Rename DECEMBER/JANUARY Column
December_sleep<-December_sleep%>%rename(DayOfTheWeek=DECEMBER)
## Add Flag for month
December_sleep<-December_sleep%>%mutate(
  Month="December"
)

## Do same for Januyar
January_sleep<-January_sleep%>%rename(DayOfTheWeek=JANUARY)
January_sleep<-January_sleep%>%mutate(
  Month="January"
)

## Combine
dec_jan<-December_sleep%>%bind_rows(January_sleep)
```

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

### Questions 1.2

Repeat the cleaning steps used in NOVEMBER, but adapt them for your `dec_jan` object.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

### Questions 1.3

Print out out the `dec_jan` object using the `kbl()` function.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>


### Questions 1.4

Plot the `dec_jan` object similar to example analysis on NOVEMBER.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

### Questions 1.4

Perform the exploratory analysis on `dec_jan` object similar to example analysis on NOVEMBER.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>


### Questions 1.5

Interpret your plots and analysis here:

<span style="color: red;">YOUR SOLUTION HERE</span>



<span style="color: red;">END OF YOUR SOLUTION HERE</span>

