---
title: 'Lab 3 Inclass: Google Location History'
author: '[PUT YOUR NAME HERE]'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
subtitle: 'SOC 325: Quantified-Self'
---

## Preamble before starting the labs

All of the R and RMarkdown files of the lab files for this class are in [github](https://github.com/zalmquist/soc325Rlab). However, PythonHub will have them all pre-installed.


# Getting started: Editing and running code

There are several ways we can run code:

  - Highlight lines in the editor window and click **Run** at the top or hit `Ctrl` + `Enter` or `CMD` + `Enter` (Mac) to run them all.
  - With your caret on a line you want to run, hit `Ctrl` + `Enter` or `CMD` + `Enter` (Mac). Note your caret moves to the next line, so you can run code sequentially with repeated presses. 
  - Type individual lines in the console and press enter. 
  - In R Markdown documents, click within a code chunk and click the green arrow to run the chunk. The button beside that runs all prior chunks. 
  

# Load Libraries Needed For this Lab

```{r,message=FALSE,warning=FALSE}
## Run this code, to manage packages and install as needed.
# Install pacman
if (!require("pacman")) install.packages("pacman")
# p_load function loads packages if installed, or install then loads otherwise
pacman::p_load(jsonlite, ggmap,ggplot2, lubridate, zoo,here,tidyverse,kableExtra,ggmap)
```

# Data

Load example data from Google Location History from [Kaggle](https://www.kaggle.com/datasets/juliansimon/location-history?resource=download) provided by the google data downloand (takeout)[https://takeout.google.com/settings/takeout/custom/location_history].

```{r,cache=TRUE}
data <- fromJSON(here("data","location_history_102014.json"))
head(data[[1]])%>%kableExtra::kable()
```

## Clean Data

```{r}
locations = data[[1]]

## The coordinates are stored in E7 format and should be converted to regular GPS coordinates.
locations$lat = locations$latitudeE7 / 1e7
locations$lon = locations$longitudeE7 / 1e7

## Fix Date
locations$time = as.POSIXct(as.numeric(locations$timestampMs)/1000, origin = "1970-01-01")
locations$date <- as.Date(locations$time, '%Y/%m/%d',tz="America/Los_Angeles")
locations$year <- year(locations$date)
locations$month_year <- as.yearmon(locations$date)
```

# How often is data recorded?

```{r}
## Count points per day, month and year
points_p_day <- data.frame(table(locations$date), group = "day")
points_p_month <- data.frame(table(locations$month_year ), group = "month")
points_p_year <- data.frame(table(locations$year), group = "year")

## Check Points per day
head(points_p_day)
summary(points_p_day$Freq)

head(points_p_month)
summary(points_p_month$Freq)

head(points_p_year)
summary(points_p_year$Freq)
```

## Plot Count of Data Points

```{r}
my_theme <- function(base_size = 12, base_family = "sans"){
  theme_grey(base_size = base_size, base_family = base_family) +
    theme(
      axis.text = element_text(size = 12),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      axis.title = element_text(size = 14),
      panel.grid.major = element_line(color = "grey"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "aliceblue"),
      strip.background = element_rect(fill = "lightgrey", color = "grey", size = 1),
      strip.text = element_text(face = "bold", size = 12, color = "navy"),
      legend.position = "right",
      legend.background = element_blank(),
      panel.margin = unit(.5, "lines"),
      panel.border = element_rect(color = "grey", fill = NA, size = 0.5)
    )
}

points <- rbind(points_p_day[, -1], points_p_month[, -1], points_p_year[, -1])
ggplot(points, aes(x = group, y = Freq)) + 
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) + 
  geom_boxplot(aes(color = group), size = 1, outlier.colour = NA) + 
  facet_grid(group ~ ., scales = "free") + my_theme() +
  theme(
    legend.position = "none",
    strip.placement = "outside",
    strip.background = element_blank(),
    strip.text = element_blank(),
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5)
  ) +
  labs(
    x = "",
    y = "Number of data points",
    title = "How many data points did Google collect about me?",
    subtitle = "Number of data points per day, month and year",
    caption = "\nGoogle collected between 0 and 1500 data points per day
    (median ~1400), between 0 and 45,000 per month (median ~42,000) and 
    between 56,000 and 170,000 per year (median ~115,000)."
  )
```

## Accuracy

```{r}
accuracy <- data.frame(accuracy = locations$accuracy, group = ifelse(locations$accuracy < 800, "high", ifelse(locations$accuracy < 5000, "middle", "low")))
accuracy$group <- factor(accuracy$group, levels = c("high", "middle", "low"))
ggplot(accuracy, aes(x = accuracy, fill = group)) + 
  geom_histogram() + 
  facet_grid(group ~ ., scales="free") + 
  my_theme() +
  theme(
    legend.position = "none",
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5)
  ) +
  labs(
    x = "Accuracy in metres",
    y = "Count",
    title = "How accurate is the location data?",
    subtitle = "Histogram of accuracy of location points",
    caption = "\nMost data points are pretty accurate, 
but there are still many data points with a high inaccuracy.
    These were probably from areas with bad satellite reception."
  )
```

# Map

## World Map


> get_googlemap() queries the Google Maps Static API version 2 to download a static map. Note that in most cases by using this function you are agreeing to the Google Maps API Terms of Service at https://cloud.google.com/maps-platform/terms/. Note that as of mid-2018, registering with Google Cloud to obtain an API key is required to use any of Google's services, including get_googlemap(). Usage and billing may apply, see the links under See Also further down in this documentation for more details.

```{r,eval=FALSE}
### This code wont run
### Turn eval=TRUE to run
#### Need a googel API key to use this package
GGMAP_GOOGLE_API_KEY<- "[Your Key Here]"
register_google(key = GGMAP_GOOGLE_API_KEY)

## World
world<-get_googlemap(center = c(lon = 0, lat = 0)
              , zoom = 1
              , maptype="roadmap"
              , scale = 2
              , color = "bw") 


#https://www.latlong.net/place/plaza-puerta-del-sol-madrid-spain-23211.html
spain<-get_googlemap(center = c(lon = -3.703790, lat = 40.416775)
              , zoom = 5
              , maptype="roadmap"
              , scale = 2
              , color = "bw") 

spain_zoom<-get_googlemap(center = c(lon = -3.703790, lat = 40.416775)
              , zoom = 8
              , maptype="roadmap"
              , scale = 2
              , color = "bw") 

spain_zoom2<-get_googlemap(center = c(lon = -3.703790, lat = 40.416775)
              , zoom = 10
              , maptype="roadmap"
              , scale = 2
              , color = "bw") 

```

```{r}
#####################
## Simple Map
#####################
load(here("data","maps.Rda"))

world%>% ggmap()+ geom_point(data = locations, aes(x = lon, y = lat), alpha = 0.5, color = "red") + 
  theme(legend.position = "right") + 
  labs(
    x = "Longitude", 
    y = "Latitude", 
    title = "Location history data points in the world",
    caption = "\nA simple point plot shows recorded positions.")
```


## Spain, European Union


```{r}

spain%>%ggmap()+geom_point(data = locations, aes(x = lon, y = lat), alpha = 0.5, color = "red") + 
  theme(legend.position = "right") + 
  labs(
    x = "Longitude", 
    y = "Latitude", 
    title = "Location history data points in Spain",
    caption = "\nA simple point plot shows recorded positions.")

spain_zoom%>%ggmap()+geom_point(data = locations, aes(x = lon, y = lat), alpha = 0.5, color = "red") + 
  theme(legend.position = "right") + 
  labs(
    x = "Longitude", 
    y = "Latitude", 
    title = "Location history data points in Spain",
    caption = "\nA simple point plot shows recorded positions.")

spain_zoom2%>%ggmap()+geom_point(data = locations, aes(x = lon, y = lat), alpha = 0.5, color = "red") + 
  theme(legend.position = "right") + 
  labs(
    x = "Longitude", 
    y = "Latitude", 
    title = "Location history data points in Spain",
    caption = "\nA simple point plot shows recorded positions.")

```



```{r}
## Loads spain_zoom3
load(here("data","maps.Rda"))

## Plot a closer zoom of Madrid here
### See code above for how to do this

```


# Spotify!

This does not work on JupyterHub unfortunately. You have to do this on R and Rstudio on your home computer. I will review it in class, but the lab exercise is extra credit! It is a bit of work to get up and running, but it can be fun to play with your spotify data.

## Necessary Packages

```{r,message=FALSE,warning=FALSE}
## Run this code, to manage packages and install as needed.
# Install pacman
if (!require("pacman")) install.packages("pacman")
# p_load function loads packages if installed, or install then loads otherwise
pacman::p_load(jsonlite, ggmap,ggplot2, lubridate, zoo,here,tidyverse,kableExtra,ggmap,spotifyr,ggjoy)
```


## Spotify API Access

* See this tutorial for how to acces the API - [spotify R API](https://www.rcharlie.com/spotifyr/).
  + Go to [https://developer.spotify.com/dashboard/](https://developer.spotify.com/dashboard/)
  + Log In and accept terms and conditions
  + Create a new app and follow directions on [spotify R API](https://www.rcharlie.com/spotifyr/)


```{r,eval=FALSE}
Sys.setenv(SPOTIFY_CLIENT_ID = 'YOURKEY')
Sys.setenv(SPOTIFY_CLIENT_SECRET = 'YOURKEY')
access_token <- get_spotify_access_token()
```

```{r,eval=FALSE}
########
## If you run the above code and this code
## You wont save your id in Markdown
########
grp<-get_my_recently_played(limit = 5) 
gta_st<-get_my_top_artists_or_tracks(type = 'artists', time_range = 'short_term', limit = 5)
gta_lt<-get_my_top_artists_or_tracks(type = 'artists', time_range = 'long_term', limit = 5)
gta_st_tracks<-get_my_top_artists_or_tracks(type = 'tracks', time_range = 'short_term', limit = 5)
gta_lt_tracks<-get_my_top_artists_or_tracks(type = 'tracks', time_range = 'long_term', limit = 5)

ed <- spotifyr::get_artist_audio_features('ed sheeran')

save(grp,gta_st,gta_lt,gta_st_tracks,gta_lt_tracks, ed,file="data/spotify.Rda")
```

```{r}
################
## Code Examples
################
## Load example data
load(here("data","spotify.Rda"))


grp%>% 
  mutate(artist.name = map_chr(track.artists, function(x) x$name[1]),
         played_at = as_datetime(played_at)) %>% 
  select(track.name, artist.name, track.album.name, played_at) %>% 
  kable()

grp %>% 
  mutate(artist.name = map_chr(track.artists, function(x) x$name[1]),
         played_at = as_datetime(played_at)) %>% 
  select(track.name, artist.name, track.album.name, played_at) %>% 
  kable()

gta_st %>% 
  mutate(artist.name = map_chr(artists, function(x) x$name[1])) %>% 
  select(name, artist.name, album.name) %>% 
  kable()

gta_lt%>% 
  select(name, genres) %>% 
  rowwise %>% 
  mutate(genres = paste(genres, collapse = ', ')) %>% 
  ungroup %>% 
  kable()


gta_st_tracks %>% 
  mutate(artist.name = map_chr(artists, function(x) x$name[1])) %>% 
  select(name, artist.name, album.name) %>% 
  kable()

ed %>%arrange(-valence) %>% 
  select(track_name, valence) %>% 
  distinct() %>%
  head(5) %>% 
  kable()

library(ggjoy)

ggplot(ed, aes(x = valence, y = album_name)) + 
  geom_joy() + 
  theme_joy() +
  ggtitle("Joyplot of Ed's joy distributions", subtitle = "Based on valence pulled from Spotify's Web API with spotifyr")
```


