---
title: "SOC 325 Lab 4 - Analyzing the Sleep Survey"
subtitle: "Soc 325: Quantified-Self"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

# Preamble before starting the labs

You'll download all R and RMarkdown files of the lab files for this class as `.zip` files from [Google Drive ](). We will use library `googlesheets4` to access data for this lab. This will allow us to limit the data access to just UW or just this class as needed.

# Pyschology of R, RStudio and Labs

Scripting is HARD. That is ok, many hard things are useful! This is very helpful if you want to work as a data scientist. You need to familiarize yourself with the practices of coding and building analysis scripts.

**Why**

* Your work has to be reproducible 
* Your work has to be able to hand off to someone else
* You need to be able to do lots of different things

**You Don't need to do everything from scratch!!**

* It is ok to borrow code, use [stack overflow](https://stackoverflow.com/), [google](https://google.com), etc.
* Use books like [R for Data Science](https://r4ds.had.co.nz/introduction.html).

**Strategies for working in RMarkdown**

* DO **NOT** code a bunch and then *knit* -- This will almost assuradely not work and crash!
* Do make small changes and *knit*
  + If you have a part that takes a long time (e.g. reading in data) turn `cache=TRUE` on in the R Chunk. This will pre-load it and make it faster.


# ATUS Sleep Score and Review of R Code

## Load Libraries Needed For this Lab

For library management we use use the library `pacman`. The key with pacman is it will install and load the packages if needed. This makes our life easier!

The first line of code is `if (!require("pacman")) install.packages("pacman")` to install pacman if needed.

The second line `pacman::p_load([PACKAGES HERE])` uses `pacman` to load and install all the packages for the lab.


```{r,message=FALSE,warning=FALSE}
## Run this code, to manage packages and install as needed.
# Install pacman
if (!require("pacman")) install.packages("pacman", repos = "http://cran.us.r-project.org")
# p_load function loads packages if installed, or install then loads otherwise
pacman::p_load(tidyverse,googlesheets4,haven,lubridate,ggthemes,AMR,kableExtra,here)
```

## Review the R Coded Needed for this lab

The first part of this lab is to review and introduce all the key R and tidyverse functions we need for the lab.


## Data Import

In this section we are going to get experience pulling from Google Sheets. To do this we will use the `googlesheets4` package load the data from Google sheets.

* This process takes a little bit of time.
* It will make you sign into your UW Gmail account.
* `cache=TRUE` is turned on so you don't have to download the data again and again.

We will use the `read_sheet([URL TO SHEET HERE])` to get the sheet information. Run this code for the first time directly in the *Console* to work through the sign in process. 

```{r,cache=TRUE,warning=TRUE,message=FALSE}
atus<-read_sheet("https://docs.google.com/spreadsheets/d/1SlgdikWNHnboJJOOs9RM0s0_VvM5NFu9_9ftcnPsjGU/edit?usp=sharing")
atus
```

## Data Cleaning and munging

First thing after we load the data is we need to clean it!

* Clean `SEX` so that 1 is Male and 2 is Female.
* Create five year age groups. To do this we will use the `AMR` package function `age_groups`, `AMR::age_groups(x,"fives")`.

To do this we will use the tidyverse format:

Object %>% verb

In this case

`atus%>%mutate([ACTION HERE])`

```{r}
## Create a new variable sex, where we rcode 1 to Male and 2 to Female
## Create a new variable age_group = age_groups(AGE,"fives") five year 
## age groups
atus<-atus%>%mutate(
  sex = if_else(SEX==1,"Male","Female"),
  age_group = age_groups(AGE,"fives")
)

atus
```


### Let's look at Seattle, SF, LA, NY, Boston, DC Metros sleep pattern

* In the ATUS `METAREA` has codes for [US Metropolitan statistical areas](https://en.wikipedia.org/wiki/Metropolitan_statistical_area).

To do this we will filter Seattle, SF, LA, NY, Boston, DC Metros and then we we will code the numeric metro names to readable metro names (e.g. Seattle Metro).

Again, we will use the object pipe verb formate (object%>%verb). In this case

`atus%>%filter([FILTER LOGICAL EXPRESSION]`

In this case we want to `filter` to limit the rows that SF, LA, NY, Boston, DC Metros. 

* 7601 is Seattle Metro
* 7360 is SF Metro
* 4481 is LA Metro
* 5601 is NY metro
* 1121 is Boston Metro
* 8841 is DC Metro

To do this we need to keep (`TRUE`) all of these metros (`c(7601,7360,4481, 5601,1121,8841)`). We can do this with an `%in%` operator which works as `[variable] %in% [SET]`. The `%in%` operator works as follows it checks to see if thing in SET exists in each row in the VARIABLE. Let's do a small example:

```{r}
## Create example
VAR<-letters
SET<-c("z","a","c","k")
## See the logical statement
VAR%in%SET

## Veify it selects what expect
### Notice order is in the VARIABLE object
letters[VAR%in%SET]
```

Let's do this now for our ATUS data,


```{r}
atus_metro<-atus%>%filter(METAREA%in%c(7601,7360,4481, 5601,1121,8841))
```

To MAP the NUMERIC (e.g. 7601) to STRING (e.g. "Seattle Metro") we will use the `case_when` function, that works as follows:

```
case_when(
    METAREA==7601~"Seattle Metro",
    METAREA==7360~"SF Metro",
    METAREA==4481~"LA Metro", 
    METAREA==5601~"NY Metro",
    METAREA==1121~"Boston Metro",
    METAREA==8841~"DC Metro"
  )
```

Now, lets put all this together, `object%>%verb` `atus_metro%>%mutate(case_when))` format:


```{r}
atus_metro<-atus_metro%>%mutate(
  location = case_when(
    METAREA==7601~"Seattle Metro",
    METAREA==7360~"SF Metro",
    METAREA==4481~"LA Metro", 
    METAREA==5601~"NY Metro",
    METAREA==1121~"Boston Metro",
    METAREA==8841~"DC Metro"
  )
)

## Print out to see if it worked
atus_metro
```

## Data Summaries

The ATUS is "weighted" survey. So to make the marginals we will need to compute the weighted version. This can be done with the following equations

$$\bar{X} = \sum_{i=1}^n \frac{w_i}{\sum_{i=1}^n w_i } x_i = \frac{\sum_{i=1}^n w_i\times x_i }{\sum_{i=1}^n w_i}$$

We can do this using the tidyverse `summarise()` function and `group_by()` function. We want to see the age by location cross tab for sleep.

### Age by Location Summary

First, we want to `group_by()` `age_group` and` location` [`group_by(age_group,location)`]. Then we want to `summarise()` by Hours Slept:


```{r,warning=FALSE,message=FALSE}
sleep_summary<-atus_metro%>%group_by(age_group,location)%>%summarise(
  sleep=sum(BLS_PCARE_SLEEP*WT06),
  total=sum(WT06),
  asleep_min = sleep/total,
  asleep_hour =  asleep_min/60 ## divide by 60 to get hours
  )

## Print out see what we did
sleep_summary
```

We are going to select out (`object%>%select(var1,...,vark)`) variables and `spread` to get `location` by `age_group` with `asleep_hour`

```{r}
ct<-sleep_summary%>%
  select(location,age_group,asleep_hour)%>%
  spread(key=age_group,asleep_hour)
```

Next we are going print out the cross tabs use `kableExtra` package which has a lot of great features, if you want to play with you can find information, [here](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)

```{r}
## Cross Tab Table
ct%>%kbl(digits = 2) %>%
  kable_styling()%>%
   kable_paper("hover", full_width = F)
```


## Data Plotting


Let's plot the cross tab as matrix plot. We will use the `geom_tile()` to do this. Remember that format of `ggplot()`

```
object %>% ggplot(aes(X,Y,[fill,group,color]))+
          geom_FUNCTION()+
          theme_NAME()+
          ...[OTHER LAYERS]
```

Below is an example using the `geom_tile()`:

```{r, warning=FALSE, messages=FALSE}
## Matrix Plot
sleep_summary%>%ggplot(aes(age_group, location, fill= asleep_hour)) + 
  geom_tile()+
  theme_bw()+
  geom_text(aes(label = round(asleep_hour,2)), colour="#FFFFFF")+
  theme(legend.position="none")+
  ggtitle(" Average Hours Slept by Metro and Age")+
  xlab("Five Year Age Group")+
  ylab("MSA")
```

# Class Sleep Survey

You'll download sleep survey data from Google Drive. This will require you to be signed into your UW gmail account. Data is provided from the Spring 2023 class.

## Load Data

```{r,cache=TRUE, warning=FALSE, message=FALSE}
demographics<-read_sheet("https://docs.google.com/spreadsheets/d/1vrYgMFticTh2kwjtdfFq8mMJ2__v46iAhOl7wPj-byo/edit?usp=share_link")

chrono_type<-read_sheet("https://docs.google.com/spreadsheets/d/1Ba7DEMgMK0wTS-kR9DtPNC4GWROpL_1Zc-GXSa1fqpc/edit?usp=sharing")

daily_sleep_log<-read_sheet("https://docs.google.com/spreadsheets/d/1hcF1adwpeuJRKWZr12KDV2nMZ0a3F9kivGb25Xl6glo/edit?usp=sharing")
```

## Explore Data


### Box plot of sleep log data
```{r}
daily_sleep_log%>%mutate(
    sleep_day=
      as.factor(sleep_day)%>% ## Relevel factor so it displays Saturday to Friday
      fct_relevel(levels(as.factor(sleep_day))[c(3,4,2,6,7,5,1)])
)%>%
    ggplot(aes(x=sleep_day,y=time_slept_hour))+
    geom_boxplot()+theme_bw()
```

### Summary tables

```{r}
## Average hours slept by day of the week
daily_sleep_log%>%group_by(sleep_day)%>%summarise(ats=mean(time_slept_hour))

## Count by Race
demographics%>%group_by(race)%>%count()

## Count by gender
demographics%>%group_by(gender)%>%count()

## Feel Best
chrono_type%>%group_by(FeelBest)%>%summarise(n=n())

## Type - Morning Person, etc
chrono_type%>%group_by(type)%>%summarise(n=n())

```

### Chronotype plot of preference of sleep and wake time

```{r}
ct_long<-chrono_type%>%pivot_longer(cols=c(std_sleep_weekday:std_wake_weekday,std_sleep_weekend:std_wake_weekend), 
                  names_to="Sleep",values_to = "HourOfDay")%>%
                  mutate(
                    Sleep=str_remove(Sleep,"std_"),
                    Sleep=str_replace(Sleep,"_weekday"," Weekday"),
                    Sleep=str_remove(Sleep,"std_"),
                    Sleep=str_replace(Sleep,"_weekend"," Weekend"),
                    Sleep=str_to_title(Sleep),
                    HourOfDay=format(HourOfDay,"%H:%M:%S")
                  )%>%select(Sleep,HourOfDay)

```

```{r}
ct_long%>%
  ggplot(aes(y=HourOfDay,fill=Sleep))+
  geom_bar()+
  scale_color_fivethirtyeight() +
  theme_fivethirtyeight()+
  ylab("Hour of the Day")+
  xlab("Count")+
  guides(fill=guide_legend(title=""))
```

## Assignment

## Question 1.1

Link the demographics (`demographics`) and chrono type (`chrono_type`) data sets. *hint* use `left_join`.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

## Question 1.2

Take your linked data and create summary tables of `FeelBest` and `type` by `gender`.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>


## Question 1.3

Build a bar chart of type by gender.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>


## Question 2.1

Link the `daily_sleep_log` with your demography and chrono type linked data. *hint* use `left_join` again.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>


## Question 2.2

Link the `daily_sleep_log` with your demography and chrono type linked data. *hint* use `left_join` again.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>


## Question 2.2

Compare the Sleep time of Seattle Metro from the ATUS to our Class Average. Does it look like our class is sleeping a similar amount of time?

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>


## Question 2.3

Plot average time slept by day and gender.

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

## Question 3

Come up with a plot not thought of in this lab to illustrate a feature of the data you find interesting. 

* Step one: Pose a question (that can be answered with our survey data)
* Step two: Build your data set that can answer from our survey data
* Step three: Plot it
* Step four: Interpret your graph

<span style="color: red;">YOUR SOLUTION HERE</span>

```{r}
## Your code here
```

<span style="color: red;">END OF YOUR SOLUTION HERE</span>

